---
title: "Getting started"
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

fastreg eases the conversion of SAS register files (`.sas7bdat`) into
[Parquet](https://parquet.apache.org/). In this context, a *register*
refers to a collection of related data files that belong to the same
dataset, typically with yearly snapshots (e.g.,
`kontakter_2020.sas7bdat`, `kontakter_2021.sas7bdat`).

## Why Parquet?

[Parquet](https://parquet.apache.org/) is a columnar storage file format
optimised for analytical workloads. Compared to
[SAS](https://fileinfo.com/extension/sas7bdat) files (and row-based
formats like CSV), Parquet offers:

-   **Reduced storage**: Efficient compression significantly reduces
    disk space, especially for large datasets.
-   **Faster queries**: The columnar layout speeds up analytical queries
    that only need a subset of columns.
-   **Wide tool support**: Parquet is supported across data processing
    frameworks in [R](https://www.r-project.org/),
    [Python](https://www.python.org/), and beyond, making it easy to
    integrate into modern workflows.

## Setup

For the examples below, we simulate SAS register data for two registers,
`kontakter` and `diagnoser`:

```{r prepare}
#| code-fold: true
#| code-summary: "Show setup code"

library(fastreg)

sas_dir <- fs::path_temp("sas-dir")
kontakter_dir <- fs::path(sas_dir, "kontakter")
diagnoser_dir <- fs::path(sas_dir, "diagnoser")
fs::dir_create(kontakter_dir)
fs::dir_create(diagnoser_dir)

kontakter_list <- simulate_register(
  "kontakter",
  c("", "1999_1", "1999_2", "2020"),
  n = 1000
)
save_as_sas(kontakter_list, kontakter_dir)

diagnoser_list <- simulate_register(
  "diagnoser",
  c("2020", "2021"),
  n = 1000
)
save_as_sas(diagnoser_list, diagnoser_dir)
```

```{r setup-tree}
#| echo: false

withr::with_dir(fs::path_dir(sas_dir), {
  fs::dir_tree(fs::path_file(sas_dir))
})
```

## Converting a single file

```{r convert-file}
sas_file <- fs::path(kontakter_dir, "kontakter_2020.sas7bdat")
output_file_dir <- fs::path_temp("output-file-dir")

convert_file(
  path = sas_file,
  output_dir = output_file_dir)
```

Even though this only converts a single file, the output is partitioned by year extracted from the file name:

```{r output-tree-file}
#| echo: false

withr::with_dir(fs::path_dir(output_file_dir), {
  fs::dir_tree(fs::path_file(output_file_dir))
})
```

## Converting a single register

Use `list_sas_files()` to find SAS files in a directory (and any
subdirectories), then use `convert_register()` to convert them.
`convert_register()` expects files to be from the **same register**
based on file names.

```{r locate}
kontakter_sas_files <- list_sas_files(kontakter_dir)
kontakter_sas_files
```

```{r convert-register}
output_register_dir <- fs::path_temp("output-register-dir")

convert_register(
  path = kontakter_sas_files,
  output_dir = output_register_dir
)
```

As with `convert_file()`, the output is partitioned by year, extracted from 
file names:

```{r output-tree}
#| echo: false

withr::with_dir(fs::path_dir(output_register_dir), {
  fs::dir_tree(fs::path_file(output_register_dir))
})
```

`convert_register()` reads files in chunks (to be able to handle
larger-than-memory data), extracts 4-digit years from filenames for
partitioning, and lowercases column names. See `?convert_register` for
more details.

<!-- TODO: Maybe this isn't needed? It does cover one scenario that isn't covered above; larger than chunk size files -->

:::::: {.callout-note collapse="true" title="Conversion scenarios example"}
The diagram below shows four scenarios when converting a SAS register
files for a register called `bef`:

::::: columns
::: column
#### Input: SAS files

```
sas_dir/
├── A) bef2020.sas7bdat
├── B) bef2021.sas7bdat
├── December_2023/
│   └── B) bef2021.sas7bdat
├── C) bef2022.sas7bdat
└── D) bef.sas7bdat
```
:::

::: column
#### Output: Parquet files

```
bef/
├── year=2020/
│   └── part-c28221.parquet
├── year=2021/
│   ├── part-bf73dc.parquet
│   └── part-546bed.parquet
├── year=2022/
│   ├── part-7c041e.parquet
│   └── part-8869b7.parquet
└── year=__HIVE_DEFAULT_PARTITION__/
    └── part-a8d52c.parquet
```
:::
:::::

A)  Single file → single Parquet file, partitioned by year from
    filename.
B)  Multiple files with the same year → separate Parquet files in the
    same partition. Identical rows are not deduplicated.
C)  Large file → split into multiple Parquet files due to more rows in
    than `chunk_size`.
D)  File without year in name → placed in
    `year=__HIVE_DEFAULT_PARTITION__` (the [Apache
    Hive](https://hive.apache.org/docs/latest/user/configuration-properties/)
    default for missing partitions).
::::::

## Converting multiple registers in parallel

For many or large files, fastreg provides a
[targets](https://docs.ropensci.org/targets/) pipeline template that
parallelises conversion across CPU cores. By default it uses 10 workers,
but that can be adjusted in the pipeline to fit the available resources.

### Create the pipeline file

```{r use-targets}
# Create temp directory for example.
pipeline_dir <- fs::path_temp("pipeline-dir")
fs::dir_create(pipeline_dir)

use_targets_template(path = fs::path(pipeline_dir, "_targets.R"))
```

### Set input and output paths

Open `_targets.R` and edit the `config` section:

```{r config}
config <- list(
  input_dir = fs::path_temp("sas-dir"),
  output_dir = fs::path_temp("parquet-registers")
)
```

-   **`input_dir`**: directory containing SAS files (searched
    recursively).
-   **`output_dir`**: directory where Parquet files will be written.

### Run the pipeline

```{r tar-make}
#| eval: false
targets::tar_make()
```

```{r edit-and-run-pipeline}
#| echo: false
template_content <- readLines(fs::path(pipeline_dir, "_targets.R"))
modified_content <- template_content |>
  stringr::str_replace("/path/to/register/sas/files/directory", config$input_dir) |>
  stringr::str_replace("/path/to/output/directory",config$output_dir)

withr::with_dir(pipeline_dir, {
  writeLines(modified_content, "_targets.R")
  targets::tar_make(callr_function = NULL, reporter = "silent")
})
```

The pipeline finds all SAS files, groups them by register name, and
converts each in parallel. Re-running `tar_make()` only re-converts
registers whose source files have changed.

```{r output-pipeline}
#| echo: false
withr::with_dir(fs::path_dir(config$output_dir), {
  fs::dir_tree(fs::path_file(config$output_dir))
})
```

## Reading a Parquet register

Use `read_register()` to read converted data as a DuckDB table:

```{r read-register}
register <- read_register(output_register_dir)
register
```

You can pass a directory to read the full partitioned register or a file
path to read a single `.parquet` file. The data is read lazily, so it
won't load into memory until collected with e.g. `dplyr::collect()`.
