---
title: "Getting started"
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
#| include: false
show_tree <- function(dir) {
  withr::with_dir(fs::path_dir(dir), {
    fs::dir_tree(fs::path_file(dir))
  })
}
```

fastreg aims to make working with Danish registers simpler and faster by
providing functionality to convert the SAS register files (`.sas7bdat`)
into [Parquet](https://parquet.apache.org/) and read the resulting
Parquet files. A *register* in this context refers to a collection of
related data files that belong to the same dataset, typically with
yearly snapshots (e.g., `bef2020.sas7bdat`,`bef2021.sas7bdat`).

::: callout-note
We use package prefixes (`fastreg::`) throughout the documentation
rather than `library()` calls, to make the package origin of each
function explicit and avoid naming conflicts.
:::

## Why Parquet?

[Parquet](https://parquet.apache.org/) is a columnar storage file format
optimised for analytical workloads. Compared to
[SAS](https://fileinfo.com/extension/sas7bdat) files (and row-based
formats like CSV), Parquet offers:

- **Smaller file size**: Efficient compression significantly reduces
  disk space, especially for large datasets.
- **Faster queries**: The columnar layout speeds up analytical queries
  that only need a subset of columns.
- **Wide tool support**: Parquet is supported across data processing
  frameworks in [R](https://www.r-project.org/),
  [Python](https://www.python.org/), and beyond, making it easy to
  integrate into modern workflows.

## Setup

For the examples below, we've simulated SAS register data for two
registers, `bef` and `lmdb`:

```{r prepare}
#| code-fold: true
#| code-summary: "Show setup code"
sas_dir <- fs::path_temp("sas-dir")
fs::dir_create(sas_dir)

bef_list <- fastreg::simulate_register(
  "bef",
  c("", "1999", "1999_1", "2020"),
  n = 1000
) |>
  # Randomly (re)generate koen, so we don't rely on any potential future
  # changes to the simulated data from osdc.
  purrr::map(\(x) {
    x |> dplyr::mutate("koen" = sample(c(1, 2), 1000, replace = TRUE))
  })

lmdb_list <- fastreg::simulate_register(
  "lmdb",
  c("2020", "2021"),
  n = 1000
)

fastreg::save_as_sas(
  c(bef_list, lmdb_list),
  sas_dir
)
```

```{r setup-tree}
#| echo: false
show_tree(sas_dir)
```

## Settings to correct paths

Many of fastreg's functions depend on the locations of the original SAS
files and the eventual Parquet files including the conversion, writing,
and reading functions. Through `options()` you can set these paths in
two settings: `fastreg.project_rawdata_dir` and
`fastreg.project_workdata_dir`. You can set these `options()` at the top
of your R script or Quarto document, in your R Project's `.Rprofile`, or
in your user-level `.Rprofile`. To add to the file, at the top of an R
script, write (using a temporary directory here for these examples):

```{r options}
options(
  # With a fake project ID.
  fastreg.project_rawdata_dir = fs::path_temp("rawdata/701010/"),
  fastreg.project_workdata_dir = fs::path_temp("workdata/701010/")
)
```

If you want to set those exact same options in the R Project's
`.Rprofile`, run the following line in your Console to open up the
`.Rprofile` file for the project:

```{r options-project-profile}
#| filename: "Console"
#| eval: false
usethis::edit_r_profile("project")
```

You can then add the same `options()` as shown in the R script example
above to that file and save it. The next time you open the project,
those options will be set.

If you want to set these options for all of your R projects and
sessions, you can add them globally in your user-level `.Rprofile`. To
open the `.Rprofile`, run:

```{r options-user-profile}
#| filename: "Console"
#| eval: false
usethis::edit_r_profile("user")
```

## Converting a single file

Converting one file from SAS to Parquet in fastreg isn't a simple change
of file extension. We make use of Parquet's Hive partitioning to
organise the output by year, for easier querying and management. So the
output Parquet file is written to a subdirectory named after the year
extracted from the file name. Use the `convert()` function to convert a
single SAS file to a year-partitioned Parquet format:

```{r convert-file}
sas_file <- fs::path(sas_dir, "bef2020.sas7bdat")
output_file_dir <- fs::path_temp("output-file-dir")

fastreg::convert(
  path = sas_file,
  output_dir = output_file_dir
)
```

`convert()` reads files in chunks (to be able to handle
larger-than-memory data) with a default of reading 1 million rows,
extracts 4-digit years from filenames for partitioning, and lowercases
column names. See `?convert` for more details.

::: callout-note
When a SAS file contains more rows than the `chunk_size`, multiple
Parquet files will be created from it. This doesn't affect how the data
is loaded with `read_register()` (see
[Reading a Parquet register](#reading-a-parquet-register) below), it
only means you may see more Parquet files in the output than input SAS
files.
:::

Even though this only converts a single file, the output is partitioned
by the year extracted from the file name as seen below:

```{r output-tree-file}
#| echo: false
show_tree(output_file_dir)
```

## Converting multiple registers in parallel

For many or large files, fastreg provides a
[targets](https://docs.ropensci.org/targets/) pipeline template that
parallelises conversion across CPU cores. By default it uses 10 workers,
but that can be adjusted in the pipeline in the `_targets.R` file to not
consume too many cores on a shared server.

To create the pipeline file, you can use the `use_template()` function.
In this example, we're outputting it to a temporary directory.

```{r use-targets}
pipeline_dir <- fs::path_temp("pipeline-dir")
fs::dir_create(pipeline_dir)

fastreg::use_template(path = pipeline_dir)
```

Once the `_targets.R` file is created, open it and edit the `config`
section:

```{r config}
config <- list(
  sas_paths = fastreg::list_sas_files(fs::path_temp("sas-dir")),
  output_dir = fs::path(pipeline_dir, "parquet-registers")
)
```

The `input_dir` is the directory that contains the SAS files (searched
recursively). This directory can contain different registers. The
`output_dir` directory is where the Parquet files will be written to.

After you've updated the `config` section, you can run the pipeline:

```{r tar-make}
#| eval: false
targets::tar_make()
```

The pipeline will find all SAS files from `input_dir` and convert each
file into a Parquet file, all done in parallel. Re-running `tar_make()`
only re-converts registers whose source files have changed or if the
pipeline itself has been edited.

## Reading a Parquet register

The final function reads the converted Parquet register data into R,
returning a [DuckDB](https://duckdb.org/) table. Using a DuckDB table is
a powerful way to query and process large data without loading it all
into memory.

You can pass a directory to read a full partitioned register or a file
path to read a single Parquet file:

```{r read-file}
file <- fastreg::read_register(output_file_dir)
file
```

The resulting DuckDB table can be filtered and transformed with `dplyr`.
For example, you can filter the data:

```{r filter-file}
file |>
  dplyr::filter(koen == 2) |>
  dplyr::compute()
```

After the query (filer and mutate), we execute it with
`dplyr::compute()`. This save the result as a temporary table inside
DuckDB, without loading it into R memory.

Notice the `??` in the first line of the output. This shows us that the
total number of matching rows is not yet known because the data isn't
loaded into memory.

::: callout-note
If you need to load the data into memory in R, you can use
`dplyr::collect()`. However, for large registers this can take a long
time, so only do this when it's absolutely necessary and make sure to
filter the data before collecting.
:::
