---
title: "Getting started"
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
#| include: false
show_tree <- function(dir) {
  withr::with_dir(fs::path_dir(dir), {
    fs::dir_tree(fs::path_file(dir))
  })
}
```

One of the main purposes of fastreg is to ease the conversion of SAS
register files (`.sas7bdat`) into
[Parquet](https://parquet.apache.org/). A *register* in this context
refers to a collection of related data files that belong to the same
dataset, typically with yearly snapshots (e.g.,
`kontakter2020.sas7bdat`, `kontakter2021.sas7bdat`).

## Why Parquet?

[Parquet](https://parquet.apache.org/) is a columnar storage file format
optimised for analytical workloads. Compared to
[SAS](https://fileinfo.com/extension/sas7bdat) files (and row-based
formats like CSV), Parquet offers:

- **Smaller file size**: Efficient compression significantly reduces
    disk space, especially for large datasets.
- **Faster queries**: The columnar layout speeds up analytical queries
    that only need a subset of columns.
- **Wide tool support**: Parquet is supported across data processing
    frameworks in [R](https://www.r-project.org/),
    [Python](https://www.python.org/), and beyond, making it easy to
    integrate into modern workflows.

## Setup

For the examples below, we've simulated SAS register data for two
registers, `kontakter` and `diagnoser`:

```{r prepare}
#| code-fold: true
#| code-summary: "Show setup code"
library(fastreg)

sas_dir <- fs::path_temp("sas-dir")
fs::dir_create(sas_dir)

kontakter_list <- simulate_register(
  "kontakter",
  c("", "1999_1", "1999_2", "2020"),
  n = 1000
)

diagnoser_list <- simulate_register(
  "diagnoser",
  c("2020", "2021"),
  n = 1000
)

save_as_sas(
  c(kontakter_list, diagnoser_list),
  sas_dir
)
```

```{r setup-tree}
#| echo: false
show_tree(sas_dir)
```

## Converting a single file

Converting one file from SAS to Parquet in fastreg isn't a simple change
of file extension. We make use of Parquet's Hive partitioning to
organise the output by year, for easier querying and management. So the
output Parquet file is written to a subdirectory named after the year
extracted from the file name. You use the `convert_file()` function to
convert a single file to an output directory and the function will
create the Parquet file in the correct partition format:

```{r convert-file}
sas_file <- fs::path(sas_dir, "kontakter2020.sas7bdat")
output_file_dir <- fs::path_temp("output-file-dir")

convert_file(
  path = sas_file,
  output_dir = output_file_dir
)
```

Even though this only converts a single file, the output is partitioned
by year extracted from the file name.

```{r output-tree-file}
#| echo: false
show_tree(output_file_dir)
```

## Converting a single register

Use `list_sas_files()` to find SAS files in a directory (and any
subdirectories), then use `convert_register()` to convert them.
`convert_register()` expects files to be from the **same register**
based on file names.

```{r locate}
kontakter_sas_files <- list_sas_files(sas_dir) |>
  stringr::str_subset("kontakter")
kontakter_sas_files
```

```{r convert-register}
output_register_dir <- fs::path_temp("output-register-dir")

convert_register(
  path = kontakter_sas_files,
  output_dir = output_register_dir
)
```

As with `convert_file()`, the output is partitioned by year, extracted
from file names.

```{r output-tree}
#| echo: false
show_tree(output_register_dir)
```

`convert_register()` reads files in chunks (to be able to handle
larger-than-memory data), extracts 4-digit years from filenames for
partitioning, and lowercases column names. See `?convert_register` for
more details.

There are four different ways that the SAS files can be converted into
Parquet. Using the `bef` register as an example, if we have a set of SAS
files in the `Grunddata/` folder that looks like below:

``` text
Grunddata/
├── bef2020.sas7bdat # <1>
├── bef2021.sas7bdat # <2>
├── December_2023/
│   └── bef2021.sas7bdat # <2>
├── bef2022.sas7bdat # <3>
└── bef.sas7bdat # <4>
```

1. A single SAS file is converted to a single Parquet file, partitioned
    by year from filename.
2. Multiple SAS files with the same register and year are converted
    into separate Parquet files in the same partition (shown below).
    Rows between these several SAS files are not deduplicated, so you'll
    have to check for duplicates after conversion.
3. A large SAS file is split into multiple Parquet files that are only
    as many rows as is determined by the `chunk_size` option.
4. A SAS file without a year in file name is placed in the
    `year=__HIVE_DEFAULT_PARTITION__/` folder (the
    [Apache Hive](https://hive.apache.org/docs/latest/user/configuration-properties/)
    default for missing partitions) when it is converted to Parquet.

After converting the files into the Parquet Hive partition structure,
the SAS files from above are output into something that looks like
below:

``` text
bef/
├── year=2020/
│   └── part-c28221.parquet
├── year=2021/
│   ├── part-bf73dc.parquet
│   └── part-546bed.parquet
├── year=2022/
│   ├── part-7c041e.parquet
│   └── part-8869b7.parquet
└── year=__HIVE_DEFAULT_PARTITION__/
    └── part-a8d52c.parquet
```

## Converting multiple registers in parallel

For many or large files, fastreg provides a
[targets](https://docs.ropensci.org/targets/) pipeline template that
parallelises conversion across CPU cores. By default it uses 10 workers,
but that can be adjusted in the pipeline in the `_targets.R` file to not
consume too many cores on a shared server.

To create the pipeline file, you can use the `use_targets_template()`
function. In this example, we're outputting it to a temporary file.

```{r use-targets}
pipeline_dir <- fs::path_temp("pipeline-dir")
fs::dir_create(pipeline_dir)

use_targets_template(path = pipeline_dir)
```

Once the `_targets.R` file is created, open it and edit the `config`
section:

```{r config}
config <- list(
  input_dir = fs::path_temp("sas-dir"),
  output_dir = fs::path_temp("parquet-registers")
)
```

The `input_dir` is the directory that contains the SAS files (searched
recursively). This directory can contain many different registers,
rather than just one. The `output_dir` directory is where the Parquet
files will be written to.

After you've updated the `config` section, you can run the pipeline:

```{r tar-make}
#| eval: false
targets::tar_make()
```

```{r edit-and-run-pipeline}
#| echo: false
#| eval: !expr rlang::is_installed("targets")
template_content <- readLines(fs::path(pipeline_dir, "_targets.R"))
modified_content <- template_content |>
  stringr::str_replace("/path/to/register/sas/files/directory", config$input_dir) |>
  stringr::str_replace("/path/to/output/directory", config$output_dir)

withr::with_dir(pipeline_dir, {
  writeLines(modified_content, "_targets.R")
  targets::tar_make(callr_function = NULL, reporter = "silent")
})
```

The pipeline will find all SAS files from all the registers listed and
convert each file into a Parquet file, all done in parallel. Re-running
`tar_make()` only re-converts registers whose source files have changed
or if the pipeline itself has been edited.

Below, you can see the output of running the pipeline with the example
data:

```{r output-pipeline}
#| eval: !expr rlang::is_installed("targets")
#| echo: false
show_tree(config$output_dir)
```

## Reading a Parquet register

The final function is to read the converted Parquet register data into
R. fastreg reads the data into a [DuckDB](https://duckdb.org/) table,
which a powerful way to query and process large data.

```{r read-register}
register <- read_register(output_register_dir)
register
```

You can pass a directory to read the full partitioned register or a file
path to read a single `.parquet` file. The data is read lazily, so it
won't load into memory until collected with e.g. `dplyr::collect()`.
