---
title: "Loading the data: Importing the Parquet format using DuckDB"
vignette: >
  %\VignetteIndexEntry{Loading the data: Importing the Parquet format using DuckDB}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  eval: false
---

Now that we have the data stored in the Parquet format, we want some way
of loading it into R to use it. Here I recommend and show how to use the
SQL-variant DuckDB to work with the data. I'll explain about DuckDB, why
I'm suggesting using it, and how to use it.

## What is SQL and DuckDB?

I mention SQL briefly in the `vignette("saving-as-parquet")`, but I'll
expand more here. SQL stands for "Structured Query Language" and is a
widely used and well established way of working and interacting with
databases. There are a couple of potentially terms here that I should
clarify what they mean.

-   Database: Also called a "relational database". This specifically
    refers a file that contains to one or more "tables" of data. A table
    is a rectangular data object, like what you would see in a
    spreadsheet (rows and columns). So a database can contain multiple
    tables of data that are connected. In our case, a database of the
    registers would have the PNR (unique personal ID) that connects all
    the tables together.

-   Query: To query something means to request something from something
    else. In the case of querying a database, it means we want to get
    some data from the database. For instance, if we hypothetically want
    PNR, income, and diabetes status of those born after 2000, we would
    query the database tables for the columns PNR, income, and diabetes
    status of all persons whose birth date was after 2000.

-   Structured: In this case, this means that it is a well defined way
    of querying a database that is the most efficient. Because
    "structured" is part of SQL, this is not something you as the user
    have to think about, SQL (mostly) automatically makes the query
    efficient for you.

-   Language: Here this means that SQL is a programming language that
    you write code to do actions. Thankfully, we R users don't need to
    actually know how to write SQL because dplyr converts many of its
    commands automatically into SQL commands when the dataset we are
    working on is a SQL database. More on this later.

Ok, but what is DuckDB? Well, SQL is a generic term for any programming
language that interacts with databases. There are dozens of SQL variants
out there, with common ones being MySQL, SQLite, or Postgres. However, a
main disadvantage of these other variants is that they aren't designed
for data analysis in mind. Many of them are designed with needs that
businesses and industry people have, not researchers or data scientists.
DuckDB is a variant that is designed specifically with data analysis in
mind. Plus, DuckDB can easily interact with our Parquet data files.

## Why should you use SQL/DuckDB?

Alright, now that you know what SQL and DuckDB are, why should you use
it? Why not load in the data with `haven::read_sas()` or (if it is a csv
file) with data.table or readr? In our particular case, registers are
MASSIVE databases and loading the data into R and thus into memory can
take quite a long time. The big advantage of SQL/DuckDB is that it
doesn't load anything into memory and any computation done on the data
is done in an efficient, fast, and low-memory way. So we can do a lot of
initial data cleaning, wrangling, summarizing, and preparing in DuckDB
before finally loading it into memory to do the final statistical
analysis with R's functions. So basically, the reasons are:

-   It doesn't need to be loaded into R, so you can right away start
    doing data wrangling for your specific project.
-   Its super fast to do wrangling, joining, and basic summarizing, so
    you can quickly get to the more intensive data analysis sooner.
-   It doesn't use as much computer resources like memory, so is from a
    cost and environmental sustainability perspective better.

**But most importantly**, it is stupid easy to work with DuckDB within R
through dplyr (technically dbplyr) and I'll show you how!

## Working with the data while in R

The key packages you'll want to use are the duckdb and arrow packages.
So we can load them up. There's also a custom function within this
registers2parquet package that helps with loading the register databases.
Everyone on this project folder *should theoretically* have this package
installed, but let me know if you don't.

```{r setup}
library(fs)
library(duckdb)
library(arrow)
library(tidyverse, quietly = TRUE)
library(registers2parquet)
# devtools::load_all()
```

To see the list of registers available in the Parquet database, use this
function:

```{r}
list_databases()
```

Let's take a simple register as an example. The "ftforael" will do. Use
the custom `load_database()` function from this registers2parquet package.
After that, we need assign

```{r}
ftf_db <- load_database("ftforael")
ftf_db
```

You'll see that the printed output of this object says
`# Source: table<arrow_002> [?? x 12]`, or something similar. What this
means is that it is showing you a small output of an arrow table (the
Parquet data format), that it has 12 columns and an unknown number of
rows. Why unknown? Because DuckDB doesn't store anything on memory and
because of the way it processes the query to the database, it can't
actually tell you how many rows are in the database unless you
specifically ask for it. For us, we don't really need to know at this
point.

What if you want to do some wrangling. You can use almost all dplyr
functions and when you end the pipe with `compute()`, it sends the query
to the database to get processed. You don't have to end with
`compute()`, but its good practice to use it to indicate where in the
pipeline you want the query to actually get sent.

```{r}
ftf_db |>
  count(year) |>
  compute()
```
