---
title: Conversion log
format:
   pdf:
      toc: true
execute:
  echo: false
  message: false
---

## About this log

For each register, this log includes two sections:

1. Read and converted SAS files: input/output paths and row counts.
2. Schema comparisons: column names and data types are compared across
   SAS files within each register. Where schemas differ, the most common
   schema is shown alongside the differences.

If you experience any problems or bugs with `fastreg`, please add an
issue on GitHub by following this link:
<https://github.com/dp-next/fastreg/issues/new>.

```{r setup}
library(targets)
```

```{r pipeline-warnings}
#| results: asis

pipeline_warnings <- tar_meta(fields = warnings, complete_only = TRUE)

if (nrow(pipeline_warnings) > 0) {
  cat("## Warnings\n\n")
  cat(
    "The pipeline produced the following warning(s).",
    "Please review them before proceeding:\n\n"
  )
  knitr::kable(pipeline_warnings)
}
```

```{r register-log}
#| results: asis

conversion_log <- tar_read(conversion_log)

for (register in unique(conversion_log$register_name)) {
  register_log <- conversion_log |>
    dplyr::filter(register_name == register)

  cat("## ", register, "\n\n")

  register_log |>
    fastreg::print_log_row_count() |>
    fastreg::print_log_schema()
}
```
